NumPy---Array Computation for Python

### 2020-01-08 body.tex
Edit intro/history

NumPy is both a cause and an effect of a striking shift in the nature of
scientific computing.  In the traditional scientific software model, the user is a {\it consumer}.  Writing software is a task for specialists and
professionals, supported by income from selling the software to
scientists.  In the new model, the user is an {\it owner}.  The Python language, with
the NumPy library, provides a framework that scientists can consume,
but also encourages scientists to participate as programmers---to
write code for the library and the ecosystem of packages around it.  This has
fostered a new culture of contribution and collaboration in scientific
computing that resulted in a virtuous cycle of increasing use, and increasing
development.

This cycle is one of several explanations for the astonishing growth of NumPy
in scientific computing.

\section*{History}

The first incarnation of NumPy was a package called Numeric, written in 1995
by Jim Hugunin, then completing his master's thesis at MIT on
superconductor-semiconductor junctions.  Hugunin based his package on previous
work by Jim Fulton, then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.
% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
% May want to mention Yorick here (origin of broadcasting idea)

As Hugunin points out, at the time he wrote Numeric, commercial languages for
interactive computing with numerical arrays were already well-established,
including Matlab and IDL.  Back then, it must have seemed a safe bet that
Numeric would not succeed in this market.  The commercial languages were
developed and maintained by large groups of full-time professional programmers,
supported by formal quality assurance, release managers, and writers of
scientific documentation.  In contrast, Numeric, and later NumPy, would
continue to be developed almost entirely by scientists, including graduate
students and faculty, without specific funding and often to the detriment of their careers.
% S: The "Back then" sentence sets up an expectation.  Later in this section, we then explain that Numeric/NumPy *was* successful, but we never explicitly close the loop: "despite the odds, great success".

@@

% S TODO: the first sentence below could flow a bit better
Around 1998, the Space Telescope Science Institute (STScI) software group began
to use Python heavily and, around 2000, wrote a re-implementation of much of
Numeric, called NumArray, to support their need to work on large memory-mapped
arrays and arrays of mixed data type records \cite{STScI-slither}.  This
briefly caused the Numeric and NumArray communities to diverge, until 2005,
when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
This merger resulted in NumPy.  At the time, Oliphant was an assistant
professor of Electrical and Computer Engineering at Brigham Young University.

Over the next 10 years, NumPy attracted a relatively large pool of scientist
developers.  To benefit from these
user-developers, the project developed an increasingly strong culture of using
software-engineering practice to improve collaboration and reduce error
\cite{millman2014developing}. The NumPy team was early in adopting distributed
revision control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for every proposed
change to NumPy. At first, NumPy had rather sparse documentation, but the project
developed a new standard for ``docstrings''---text describing each NumPy
function---and used this standard to write comprehensive, high-quality
documentation, integrated with the source code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}. The NumPy documentation format
quickly became a standard for the larger Python community.

@@

The creation of NumPy spurred renewed development in the larger scientific
Python ecosystem and heralded the current era of wide-spread use of Python for
scientific computing.
As we implied above, there were several factors that allowed this rapid growth
and successful development.

@@

In addition, NumPy provides a large collection of utility functions for creating, reshaping, concatenating, and padding arrays; for searching, sorting and counting data in arrays; and for reading and writing files.  It provides extensive support for generating pseudorandom numbers and includes an assortment of probability distributions.  For historical reasons, NumPy also includes basic functionality for linear algebra, fast Fourier transforms and windowing, and polynomial fitting.  We will continue supporting these, but we will not expand beyond them.

>>>>

The first incarnation of NumPy was a package called Numeric, written in 1995
by Jim Hugunin, then completing his master's thesis at MIT on
superconductor-semiconductor junctions.  Hugunin based his package on previous
work by Jim Fulton, then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.
% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)

%At the time, commercial languages for interactive computing with numerical
%arrays were already well-established, including Matlab and IDL.
%They were developed and maintained by large groups of
%full-time professional programmers, supported by formal quality assurance,
%release managers, and writers of scientific documentation.
%In contrast, Numeric, and later NumPy, would continue to be developed almost
%entirely by scientists, including graduate students and faculty.
%%without specific funding and often to the detriment of their careers.
%% S: The "Back then" sentence sets up an expectation.  Later in this section, we then explain that Numeric/NumPy *was* successful, but we never explicitly close the loop: "despite the odds, great success".
%NumPy had no dedicated funding, and development time
%was mostly contributed freely by students and researchers in their
%spare time.

@@

% S TODO: the first sentence below could flow a bit better
Around 1998, the Space Telescope Science Institute (STScI) software group began
to use Python heavily and, around 2000, wrote a re-implementation of much of
Numeric, called NumArray, to support their need to work on large memory-mapped
arrays and arrays of mixed data type records \cite{STScI-slither}.  This
briefly caused the Numeric and NumArray communities to diverge, until 2005,
when Travis Oliphant embarked upon a major rewrite that aimed to be a ``best of
both worlds'' unification of Numeric and NumArray \cite{oliphant2006guide}.
This merger resulted in NumPy.

Over the next 10 years, NumPy attracted more scientist developers.
The project developed an increasingly strong culture of using
software-engineering practice to improve collaboration and reduce error
\cite{millman2014developing}. The NumPy team was early in adopting distributed
revision control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for every proposed
change to NumPy. At first, NumPy had rather sparse documentation, but the project
developed a new standard for ``docstrings''---text describing each NumPy
function---and used this standard to write comprehensive, high-quality
documentation, integrated with the source code\cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}. The NumPy documentation format
quickly became a standard for the larger Python community.

@@

The creation of NumPy spurred renewed development in the larger scientific
Python ecosystem and heralded the current era of wide-spread use of Python for
scientific computing.
There were several factors that allowed this rapid growth
and successful development.

@@

In addition, NumPy provides a large collection of utility functions for creating, reshaping, concatenating, and padding arrays; for searching, sorting and counting data in arrays; and for reading and writing files.  It provides extensive support for generating pseudorandom numbers and includes an assortment of probability distributions.
